---
title: "COVID-19 Analysis"
author: "Qua Vallery"
date: '2022-08-13'
output: pdf_document
---
# Description of Data

The COVID-19 Data Repository by the Center for Systems Science and Engineering (CSSE) at John Hopkins University includes daily COVID-19 reports for the US and globally. All data is read in from the daily case reports to generate time series summary tables including confirmed cases and deaths for the US and globally. For this analysis I will be focusing on US confirmed cases and deaths only. The two CSVs from the John Hopkins COVID-19 Data Repository used in the analysis are named time_series_covid19_confirmed_US.csv and time_series_covid19_deaths_US.csv. I developed a third CSV for this analysis named state_codes.csv. Below is a description of all columns for each dataset:

## time_series_covid19_confirmed_US.csv

Each record represents total confirmed COVID cases by  county, state between January 22, 2020, and August 5, 2022. 

- UID: Unique Identifier for each row entry.
- Iso2: Officially assigned two-letter country code identifiers.
- Iso3: Officially assigned three-letter country code identifiers.
- Code3: Officially assigned numeric country code identifiers.
- FIPS: US only. Federal Information Processing Standards code that uniquely identifies counties within the USA.
- Admin2: County name. US only.
- Province_State: Province, state or dependency name.
- Country_Region: Country, region or sovereignty name. The names of locations included on the Website correspond with the official designations used by the U.S. Department of State.
- Lat: Latitude.
- Long_: Longitude.
- Combined_Key: Combination of Admin2, Province_State, Country_Region.

## time_series_COVID19_deaths_US.csv

- UID: Unique Identifier for each row entry.
- Iso2: Officially assigned two-letter country code identifiers.
- Iso3: Officially assigned three-letter country code identifiers.
- Code3: Officially assigned numeric country code identifiers.
- FIPS: US only. Federal Information Processing Standards code that uniquely identifies counties within the USA.
- Admin2: County name. US only.
- Province_State: Province, state or dependency name.
- Country_Region: Country, region or sovereignty name. The names of locations included on the Website correspond with the official designations used by the U.S. Department of State.
- Lat: Latitude.
- Long_: Longitude.
- Combined_Key: Combination of Admin2, Province_State, Country_Region.
- Population: Population of the Admin2 (the county).

## state_codes.csv

- State: Name of US state.
- State Initial: US State two-letter abbreviation.

# Objectives

For this analysis I answered the following questions:

1. What are the total confirmed COVID-19 cases and deaths by state?
2. What does California's COVID-19 deaths look like over time?
3. Which states had the smallest and largest percent increase in deaths compared to 2020?
4. What is the COVID-19 deaths impact look like across different geographic areas of the US?
5. Is there a relationship between the number of COVID-19 cases and deaths?

# Import libraries and Data

I started by importing the necessary libraries and data.

```{r import libraries/data, echo = TRUE, results = "hide"}

# load packages
library(lubridate)
library(tidyverse)
library(tidyquant)
install.packages("usmap", repos = "http://cran.us.r-project.org")    
library(usmap)

# import files
US_confirmed_tbl <- read_csv("time_series_covid19_confirmed_US.csv")
US_deaths_tbl <- read_csv("time_series_covid19_deaths_US.csv")
state_codes_tbl <- readr::read_delim("state_codes.csv", delim = "|")

```
# Examine Data

I examined the data for any import issues, classification errors, and missing values. There were no import or classification errors. However, there were two columns, FIPS and Admin2, with missing values. I further examined the columns with missing values in each data set and discovered that the same observations were missing values in both the cases and deaths data sets for the same columns. Ultimately, I decided not to remove or impute the missing values because I wasn't relying on those columns for my analysis. 

```{r examine import pt 1, echo = TRUE}

# check for import errors
errors_confirmed_tbl <- readr::problems(US_confirmed_tbl)
errors_deaths_tbl <- readr::problems(US_deaths_tbl)
errors_codes_tbl <- readr::problems(state_codes_tbl)

# check for correct classification of fields

US_confirmed_tbl
US_deaths_tbl

```

```{r examine pt 2, echo = TRUE}
# NA check
findNAs_confirmed_tbl <- colSums(is.na(US_confirmed_tbl))

findNAs_deaths_tbl <- colSums(is.na(US_deaths_tbl))


```

```{r examine pt 3, echo = TRUE}

# Extract rows with NAs
    
    NAs_Confirmed_FIPS_tbl <- US_confirmed_tbl[is.na(US_confirmed_tbl$FIPS),] 
    NAs_Confirmed_Admin2_tbl <- US_confirmed_tbl[is.na(US_confirmed_tbl$Admin2),] 
    
    NAs_Deaths_FIPS_tbl <- US_deaths_tbl[is.na(US_deaths_tbl$FIPS),]
    NAs_Deaths_Admin2_tbl <- US_deaths_tbl[is.na(US_deaths_tbl$Admin2),]
    
    NAs_Combined_FIPS_tbl <- NAs_Confirmed_FIPS_tbl %>%
        left_join(y = NAs_Deaths_FIPS_tbl, by = "UID")
    
    NAs_Combined_Admin2_tbl <- NAs_Confirmed_Admin2_tbl %>%
        left_join(y = NAs_Deaths_Admin2_tbl, by = "UID")
    
    NAs_Combined_FIPS_tbl
    NAs_Combined_Admin2_tbl

```

# Tidy and Transform


```{r tiday and transform, echo = TRUE}

# select columns needed for analysis
    
    US_confirmed_stripped_tbl <- US_confirmed_tbl %>%
        select(-iso2, -iso3, -code3, -FIPS)
    US_deaths_stripped_tbl <- US_deaths_tbl %>%
        select(-iso2, -iso3, -code3, -FIPS)
    
# pivot longer
    
    US_confirmed_longer_tbl <- US_confirmed_stripped_tbl %>%
        pivot_longer(
            cols = `1/22/20`: `8/5/22`,
            names_to = "Date",
            values_to = "Total Confirmed"
        )
    
    US_confirmed_longer_tbl %>% glimpse()
    
    US_deaths_longer_tbl <- US_deaths_stripped_tbl %>%
        pivot_longer(
            cols = `1/22/20`: `8/5/22`,
            names_to = "Date",
            values_to = "Total Deaths"
        )
    
    US_deaths_longer_tbl %>% glimpse()

# lubridate
    
    US_confirmed_longer_tbl$Date <- mdy(US_confirmed_longer_tbl$Date)
    US_confirmed_longer_tbl %>% glimpse()
    
    US_deaths_longer_tbl$Date <- mdy(US_deaths_longer_tbl$Date)
    US_deaths_longer_tbl %>% glimpse()
    
# groupby and summarize 
    
    US_confirmed_y_tbl <- US_confirmed_longer_tbl %>%
        mutate(Year = year(Date)) %>%
        mutate(Date = as.character(Date)) %>%
        filter(Date == "2020-12-31" | Date == "2021-12-31" | Date == "2022-08-05") %>% 
        
        group_by(UID, Admin2, Province_State, Year) %>% 
        
        summarize(Total_Confirmed = sum(`Total Confirmed`)) %>%
        ungroup()
    
    
    US_deaths_y_tbl <- US_deaths_longer_tbl%>%
        mutate(Year = year(Date)) %>%
        mutate(Date = as.character(Date)) %>%
        filter(Date == "2020-12-31" | Date == "2021-12-31" | Date == "2022-08-05") %>% 
        
        group_by(UID, Admin2, Province_State, Population, Year) %>%
        summarize(`Total Deaths` = sum(`Total Deaths`)) %>%
        ungroup()

        US_deaths_y_m_tbl <- US_deaths_longer_tbl %>%
        mutate(Year = year(Date)) %>%
        mutate(Month = month(Date, label = TRUE)) %>%
        mutate(Year_Month = ceiling_date(x = Date, unit = "month" ) - 1) %>%
        filter(Date == Year_Month) %>%
        
        
        group_by(Province_State,Year_Month) %>% 
        summarize(`Total Deaths` = sum(`Total Deaths`)) %>%
        ungroup()

# summarize total confirmed and deaths by state and year
    US_confirmed_y_tbl <- US_confirmed_y_tbl %>%
        group_by(Province_State, Year) %>%
        summarize(Total_Confirmed = sum(Total_Confirmed)) %>%
        ungroup()
   
    US_deaths_y_tbl <- US_deaths_y_tbl %>%
        group_by(Province_State, Year) %>%
        summarize(`Total Deaths` = sum(`Total Deaths`)) %>%
        ungroup()

# Calculate cases and deaths per year
    US_confirmed_y_tbl <- US_confirmed_y_tbl %>%
        group_by(Province_State) %>%
        mutate(Total_Confirmed_Lag = lag(Total_Confirmed, n = 1)) %>% 
        mutate(Total_Confirmed_Lag = case_when(
            is.na(Total_Confirmed_Lag) ~ 0,
            TRUE ~ Total_Confirmed_Lag)) %>% 
        mutate(diff = Total_Confirmed - Total_Confirmed_Lag) %>%
        ungroup()
    
    US_deaths_y_tbl <- US_deaths_y_tbl %>%
        group_by(Province_State) %>%
        mutate(Total_Deaths_Lag = lag(`Total Deaths`, n = 1)) %>% 
        mutate(Total_Deaths_Lag = case_when(
            is.na(Total_Deaths_Lag) ~ 0,
            TRUE ~ Total_Deaths_Lag)) %>% 
        mutate(diff = `Total Deaths` - Total_Deaths_Lag) %>%
        ungroup()
    
 # add unique identifier
    
    US_confirmed_y_tbl <- tibble::rowid_to_column(US_confirmed_y_tbl, "ID")
    
    US_confirmed_y_tbl %>% glimpse()
    
    US_deaths_y_tbl <- tibble::rowid_to_column(US_deaths_y_tbl, "ID")
    
    US_deaths_y_tbl %>% glimpse()

    US_combined_y_tbl <- US_confirmed_y_tbl %>%
        left_join(y = US_deaths_y_tbl, by = "ID")

# remove and rename columns
    
    US_combined_y_tbl <- US_combined_y_tbl %>%
        select(-Province_State.y, -Year.y) %>%
        rename(Year = Year.x, Cases_Per_Year = diff.x, Deaths_Per_Year = diff.y, Province_State = Province_State.x)
    

```

# Analysis and Visualization

## What are the total confirmed COVID-19 cases and deaths by state?

The top 6 states with the most COVID-19 cases are:
1. California
2. Texas
3. Florida
4. New York
5. Illinois
6. Pennsylvania

The top 6 states with the most COVID-19 deaths are:
1. California
2. Texas
3. Florida
4. New York
5. Pennsylvania
6. Georgia

Five out six states are on both lists, which suggests that states a with high number of COVID-19 cases also have a high number of deaths. Illinois didn't show up in the top 6 states with the most COVID-19 deaths, which suggests the state's COVID-19 death rate is lower than it's case rate. Unlike Illinois,Georgia is on the top six deaths list, which suggests COVID-19 death rate is higher than it's case rate.

```{r cases and deaths by state, echo = TRUE}

confirmed_by_state <- US_combined_y_tbl %>% 
        select(Province_State, Cases_Per_Year) %>% 
        filter(Province_State != "Diamond Princess", Province_State != "Grand Princess") %>%
        group_by(Province_State) %>%
        summarize(Total_Cases = sum(Cases_Per_Year)) %>%
        arrange(desc(Total_Cases)) %>%
        ungroup()
    
    confirmed_by_state
    
    deaths_by_state <- US_combined_y_tbl %>%
        select(Province_State, Deaths_Per_Year) %>%
        filter(Province_State != "Diamond Princess", Province_State != "Grand Princess") %>% #remove ships
        group_by(Province_State) %>%
        summarize(`Total Deaths`= sum(Deaths_Per_Year)) %>%
        arrange(desc(`Total Deaths`)) %>%
        ungroup()
    
    deaths_by_state
    
confirmed_by_state %>% head()
deaths_by_state %>% head()
    


```

## 2. What does California's COVID-19 deaths look like over time?

Because I live in California, I decided to graph California's COVID-19 deaths over time. COVID-19 deaths are only through July 2022 due to not having an entire month's worth of August COVID-19 data at the time of this analysis.  

By the end of 2020, California had about 25 thousand COVID-19 deaths. However, that number quickly doubled. By the end of the first quarter of 2021, California had surpassed 50 thousand COVID-19 deaths. Deaths slowed down in the summer of 2021 and reached 75 thousand by the end of 2021. As of July 2022, California has had approximately 93 thousand COVID-19 deaths.

```{r CA deaths over time, echo = TRUE}

# Plot California deaths over time
    

    CA_deaths_plot <- US_deaths_y_m_tbl %>%
        
        filter(Province_State == "California") %>%
        select(Year_Month, `Total Deaths`) %>%
      
        ggplot(aes(x = Year_Month, y = `Total Deaths`)) +
        geom_line(color = "cyan3", size = 1.5) +
        labs(
            title = "California COVID-19 Deaths Over Time",
            subtitle = "Jan 2020 - July 2022",
            x = "Year",
        ) +
        ylim(0, 100000) +
        theme_tq()
    
    CA_deaths_plot

```

## 3. Which states had the smallest and largest percent increase in COVID-19 deaths compared to 2020?

The top 6 states with the largest percent increase in COVID-19 deaths compared to 2020 are:
1. Northern Mariana Islands
2. Maine
3. Oklahoma
4. Alaska
5. Kentucky
6. Oregon

The top 6 states with the smallest percent increase in COVID-19 deaths compared to 2020 are:
1. Rhode Island
2. Connecticut
3. New York
4. New Jersey
5. North Dakota
6. District of Columbia

It's worth noting that the percent changes are relative. Relative changes on small numbers can appear to be more significant than they are. For example, Northern Mariana Islands had 2 COVID-19 deaths reported in 2020 and 38 reported in 2022, resulting in an 1,800 percent increase. However, the absolute number of COVID-19 deaths in Northern Mariana Islands is extremely small.

```{r measuring change, echo = TRUE}

# Calculate percent difference 
    state_pct_diff_tbl <- US_combined_y_tbl %>%
        select(Province_State, Year, `Total Deaths`) %>%
        group_by(Province_State) %>%
        mutate(year_2020 = first(`Total Deaths`)) %>% 
        mutate(diff = `Total Deaths` - year_2020) %>% 
        mutate(pct_diff = diff / year_2020) %>% #View()
        mutate(pct_diff = case_when(
            pct_diff == "NaN" ~ 0,
            pct_diff == "Inf" ~ 1,
            TRUE ~ pct_diff)) %>%
        mutate(pct_diff = round(pct_diff, 2)) %>%
        mutate(pct_diff_chr = scales::percent(pct_diff)) %>% #I need to remove this field
        ungroup()
    
    state_pct_diff_tbl %>% glimpse()
    
# Show percent difference in deaths compared to 2020    
    state_max_pct_increase <- state_pct_diff_tbl %>%
        filter(Year == "2022") %>%
        filter(Province_State != "Diamond Princess", Province_State != "Grand Princess") %>%
        group_by(Province_State) %>%
        summarize(max_increase = max(pct_diff)) %>%
        arrange(desc(max_increase)) %>% glimpse() %>%
        mutate(max_increase = scales::percent(max_increase)) %>%
        ungroup()
        
    state_max_pct_increase 
   
    largest_pct_increase <- state_max_pct_increase %>% head()
    smallest_pct_increase <- state_max_pct_increase %>% tail()
    
    largest_pct_increase
    smallest_pct_increase
    

```

## 4. What is the COVID-19 deaths impact look like across different geographic areas of the US?

There about four states, California, Texas, New York, Florida in the highest threshold with around 75 thousand deaths or more. While I know there are a lot more people in California and New York, it is still interesting to see the two states in the highest band because I do know they have some of the highest vaccination rates. It would be worth exploring when most of the deaths in California and New York happened. For example, were there were a lot of deaths early on before vaccines were available. 

```{r deaths map, echo = TRUE}

#5.4 Create a map
    
    # reorder tibble
    deaths_by_state <- deaths_by_state %>%
        arrange(Province_State)
    
    # create state codes tbl
    state_codes_tbl <- readr::read_delim("state_codes.csv", delim = "|") # I created this CSV file
    
    state_codes_tbl <- state_codes_tbl %>%
        rename(Province_State = State)
    
     # join tables
    deaths_by_state <- deaths_by_state %>%
        left_join(y = state_codes_tbl, by ="Province_State" ) %>%
        rename(values = `Total Deaths`)
    
    deaths_by_state_map_tbl <- deaths_by_state %>%
        select(-Province_State)
    
    density_map_tbl <- map_with_data(data = deaths_by_state_map_tbl, values = "values")
    density_map_tbl <- density_map_tbl %>%
        select(state, values)
    
    # Plot map
    plot_usmap(regions = "states", data = density_map_tbl, values = "values") +
    labs(title = "US COVID-19 Deaths Heat Map",
        subtitle = "Jaunary 2020 - August 2022") +
    theme(panel.background = element_blank()) +
    scale_fill_continuous(low = "yellow", high = "red", name = "Total Deaths")

```

## 5. Is there a relationship between the number of COVID-19 cases and deaths?

It appears that there is a strong relationship between the number of COVID-19 cases and deaths. Additionally, the goodness of fit measures for linear regression model, r squared, has a  value of 0.91.Meaning that the linear regression model explains the observed data well. Lastly, the p-value is significantly small at 2.2e-16, which suggests that the predictor variable, number of COVID-19 cases are associated with the response variable, number of COVID-19 deaths

```{r deaths and cases relationship, echo = TRUE}

 # update combined tbl with cases per thousand and deaths per thousand columns
    
    US_combined_y_tbl <-  US_combined_y_tbl %>%
        mutate(cases_per_thou = Total_Confirmed / 1000) %>%
        mutate(deaths_per_thou = `Total Deaths` / 1000)

    
    # estimate linear regression
    my_mod <- lm(deaths_per_thou ~ cases_per_thou, US_combined_y_tbl)
    summary(my_mod)
    
    # create data for ggplot
    data_mod_tbl <- tibble(predicted = predict(my_mod),
                           observed = US_combined_y_tbl$deaths_per_thou)
    
    # create plot
    ggplot(data_mod_tbl,
           aes(x = predicted, 
               y = observed)) +
        geom_point() +
        geom_abline(intercept = 0,
                    slope = 1,
                    color = "red",
                    size = 2) +
        labs(
            x = "Cases per Thousand",
            y = "Deaths per Thousand"
        )
        theme_tq() 

```
# Conclusion and Bias Identification

It's interesting to see the number of COVID-19 cases and deaths and relationships in the US by state as well as the death trend in California. Additional analysis could be done to look at more granular locational impacts, COVID-19 cases and deaths compared to state populations, and case and deaths rates over time.

A bias that came to mine as I was completing this project was vaccination rates. However, I didn't merge COVID-19 vaccination data to explore this biased thought.
